---
title: "Analysis of BLAST Results"
author: "Emre Ovet"
date: "October 12, 2018"
output: github_document
---

# Introduction

The human skin contains a large amount of bacteria on its surface due to its constant contact with air and different objects in daily life. Those bacterial communities are able to survive on skin surface for long periods of time because most of them are highly resistant to environmental stress factors like temperature and UV, thus, skin surface is a favorable habitat for them. 

According to the recent studies, it is found out that the diversity of bacterial communities on palm surface is much higher than what is previously assumed and that after washing hands, the  surviving bacteria are able to reproduce rapidly and recover their previous population. It is also found out that the genera of bacteria living on the palm surface largely varies between individuals.

In the light of these information, we hypothesize that the specificity of the bacterial community genera in male and female palm surfaces can be used for forensic investigation in terms of gender identification by comparing the bacteria found on different objects with those found on the palm surfaces of individuals. In order to test our hypothesis, we made several plots using the study of Fierer et al. in which bacterial communities from keyboards and computer mouses were extracted along with those from the palm surfaces of individuals using those equipment, then the genera of the bacteria from keyboards and computer mouses were compared with those from skin surfaces using pyrosequencing. The first figure will demonstrate the most abundant taxa in male vs female hands to see the level of similaries and differences between bacterial community genera in male and female palm surfaces, the second and figures will demonstrate 
# Methods

## Sample origin and sequencing

For the computer mouse analysis, the palm surfaces of 5 men and 4 women from University of Colorado campus were swabbed along with the computer mouses of each individual's personal computer which were last touched by their owners 12 hours before the swabbing. All swabs were stored in -80°C until the DNA extraction. Then, the database of bacterial communities from both the right and left hand palm surfaces of 270 other individuals were compiled, sampled and analyzed.

For the pyrosequencing, DNA was extracted from samples by using MO BIO Powersoil DNA isolation kit. Amplicons were cleaned using the UltraClean-htp 96-well PCR Clean-up kit and the DNA concentration inside the amplicon were measured using the Quant-iT PicoGreen dsDNA reagent and kit. 

## Computational

On a tmux session, QC reports were created using the data from NCBI using FastQC program. Then, we used the Cyberduck program to transfer the files from tmux to our computer. After then, we trimed the sequences by looking at their quality scores, that is, any sequence below 150 base pairs were filtered. Furthermore, we  Blasted the sequences to obtain the top match of each sequence.Finally, we used dplyr functions to visualize our data on Rstudio by making plots.



# Results

```{r load-libraries, message = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the install.packages() function
# or with the 'Packages' pane in RStudio

# load packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
```

```{r make-read-in-data-function}
# Output format from BLAST is as detailed on:
# https://www.ncbi.nlm.nih.gov/books/NBK279675/
# In this case, we used: '10 sscinames std'
# 10 means csv format
# sscinames means unique Subject Scientific Name(s), separated by a ';'
# std means the standard set of result columns, which are:
# 'qseqid sseqid pident length mismatch
# gapopen qstart qend sstart send evalue bitscore',


# this function takes as input a quoted path to a BLAST result file
# and produces as output a dataframe with proper column headers
# and the 'qseqid' column split into sample and seq number
read_blast_output <- function(filename) {
  data_in <- read.csv(filename,
                      header = FALSE, # files don't have column names in them
                      col.names = c("sscinames", # unique Subject Sci Name(s)
                                    "qseqid",    # Query Seq-id
                                    "sseqid",    # Subject Seq-id
                                    "pident",    # Percntge of identical matches
                                    "length",    # Alignment length
                                    "mismatch",  # Number of mismatches
                                    "gapopen",   # Number of gap openings
                                    "qstart",    # Start of alignment in query
                                    "qend",      # End of alignment in query
                                    "sstart",    # Start of alignment in subj
                                    "send",      # End of alignment in subject
                                    "evalue",    # Expect value
                                    "bitscore"))  # Bit score

  # Next we want to split the query sequence ID into
  # Sample and Number components so we can group by sample
  # They originally look like "ERR1942280.1"
  # and we want to split that into two columns: "ERR1942280" and "1"
  # we can use the separate() function from the tidyr library to do this
  # Note that we have to double escape the period for this to work
  # the syntax is
  # separate(column_to_separate,
  # c("New_column_name_1", "New_column_name_2"),
  # "seperator")
  data_in <- data_in %>%
    separate(qseqid, c("sample_name", "sample_number"), "\\.")
}
```

```{r read-in-BLAST-data}
# this makes a vector of all the BLAST output file names, including
# the name(s) of the directories they are in
files_to_read_in <- list.files(path = "output/blast",
                               full.names = TRUE)

# We need to create an empty matrix with the right number of columns
# so that we can rbind() each dataset on to it
joined_blast_data <- matrix(nrow = 0,
                            ncol = 14)

# now we loop over each of the files in the list and append them
# to the bottom of the 'joined_blast_data' object
# we do this with the rbind() function and the function we
# made earlier to read in the files, read_blast_output()
for (filename in files_to_read_in) {
  joined_blast_data <- rbind(joined_blast_data,
                             read_blast_output(filename))
}
```

```{r read-in-metadata-and-join}
# Next we want to read in the metadata file so we can add that in too
# This is not a csv file, so we have to use a slightly different syntax
# here the `sep = "\t"` tells the function that the data are tab-delimited
# and the `stringsAsFactors = FALSE` tells it not to assume that things are
# categorical variables
metadata_in <- read.table(paste0("data/metadata/",
                                 "fierer_forensic_hand_mouse_SraRunTable.txt"),
                          sep = "\t",
                          header = TRUE,
                          stringsAsFactors = FALSE)

# Finally we use the left_join() function from dplyr to merge or 'join' the
# combined data and metadata into one big table, so it's easier to work with
# in R the `by = c("Run_s" = "sample_name")` syntax tells R which columns
# to match up when joining the datasets together
joined_blast_data_metadata <- metadata_in %>%
  left_join(joined_blast_data,
            by = c("Run_s" = "sample_name"))
```

```{r Different bacterial taxa on male vs female hands}
joined_blast_data_metadata %>%
  group_by(sscinames, sex_s) %>%
  tally() %>%
  arrange(desc(n)) %>%
  filter(sex_s != "Not applicable") %>%
  filter(n > 100) %>%
  ggplot(aes(sscinames, y = n, fill = sex_s)) +
  geom_col() +
  theme(axis.text = element_text(angle = 90,
                                 hjust = 1)) +
  ggtitle("Number of most abundant taxa in male vs female hands")
```

*Figure 1* 
In this figure, we can see that the number of most abundant taxa in male and female hands which are higher than 100. There are only 7 bacteria out of hundreds that are widely common, and only 2 of them are found in female palm surfaces; Acidovorax sp. and Bartonella washoensis. Overall, Solemya pervemicosa gill symbiont is the most abundant bacteria in males and Bartonella washoensis is the most abundant bacteria in females.

```{r Abundance of Bartonella washoensis in dust vs sebum}
joined_blast_data_metadata %>%
  group_by(sscinames, env_material_s) %>%
  tally() %>%
  filter(sscinames %in% c("Bartonella washoensis")) %>%
  ggplot(aes(sscinames, y = n, fill = env_material_s)) +
  geom_col(position = position_dodge()) +
    theme(axis.text = element_text(angle = 90, hjust = 1)) +
  ggtitle("Abundance of Bartonella washoensis in male vs female hands")
```

*figure 2*
In this figure, we can see that the abundance of Bartonella washoensis is between 750 and 900 in dust and between 500 and 750 in sebum. 

```{r Abundance of Solemya pervernicosa gill symbiont in dust vs sebum}
joined_blast_data_metadata %>%
  group_by(sscinames, env_material_s) %>%
  tally() %>%
  filter(sscinames %in% c("Solemya pervernicosa gill symbiont")) %>%
  ggplot(aes(sscinames, y = n, fill = env_material_s)) +
  geom_col(position = position_dodge()) +
    theme(axis.text = element_text(angle = 90, hjust = 1)) +
  ggtitle("Abundance of Solemya pervernicosa gill symbiont in dust vs sebum")
```

*figure 3*
In this figure, we can see that the abundance of Solemya pervernicosa gill symbiont is between 750 and 1000 in dust and between 1500 and 1750 in sebum. 

```{r Abundance of Acidovorax sp. in dust vs sebum}
joined_blast_data_metadata %>%
  group_by(sscinames, env_material_s) %>%
  tally() %>%
  filter(sscinames %in% c("Acidovorax sp.")) %>%
  ggplot(aes(sscinames, y = n, fill = env_material_s)) +
  geom_col(position = position_dodge()) +
    theme(axis.text = element_text(angle = 90, hjust = 1)) +
  ggtitle("Abundance of Acidovorax sp. in dust vs sebum")
```

*figure 4*
In this figure, we can see that the abundance of Acidovorax sp. is between 750 and 1000 in dust and between 1500 and 1750 in sebum. 


```{r show-levels}
levels(factor(joined_blast_data_metadata$anonymized_name_s))
```


```{r Mean percent identity of query sequences}
joined_blast_data_metadata %>%
  group_by(anonymized_name_s) %>%
  summarize(mean_pident = mean(pident),
            sd_pident = sd(pident)) %>%
  ggplot(aes(x = anonymized_name_s,
             y = mean_pident)) +
  geom_col(fill = "dodger blue") +
  geom_errorbar(aes(ymax = mean_pident + sd_pident,
                    ymin = mean_pident - sd_pident),
                width = 0.3) +
theme(axis.text.x = element_text(angle = 90,
                                    hjust = 1))

```


```{r histograms}
# Here we're using the dplyr piping syntax to select a subset of rows matching a
# criteria we specify (using the filter) function, and then pull out a column
# from the data to make a histogram.
joined_blast_data_metadata %>%
  filter(env_material_s == "dust") %>%
  filter(grepl("F2", host_subject_id_s)) %>%
  ggplot(aes(x = pident)) +
    geom_histogram() +
    ggtitle("Percent Identity") +
    xlab("Percent")
```


```{r summary-table}
# Finally, we'd like to be able to make a summary table of the counts of
# sequences for each subject for both sample types. To do that we can use the
# table() function. We add the kable() function as well (from the tidyr package)
# in order to format the table nicely when the document is knitted
kable(table(joined_blast_data_metadata$host_subject_id_s,
            joined_blast_data_metadata$sample_type_s))
```

# Discussion

We found out that Solemya pervemicosa gill symbiont is the most abundant bacteria in males and Bartonella washoensis is the most abundant bacteria in females. We're wondering why this is the case but it is hard to find an answer since the information about those bacteria are very limited for now. As far as we know, Solemya pervemicosa gill symbionts are chemosynthetic bacteria, that is, they convert CH4 or CO2 into organic compounds and they provide their host with necessary nutrients. We can argue that these symbionts were somehow transferred from Solemya pervemicosa clams to humans and they had a very favorable habitat in palm surface, specifically male palm surfaces since none of them were found in female palm surfaces. One possibility is that a male individual made contact with the gills of a Solemya pervemicosa and then he made its way to the men's restroom in university campus and from here, it has spread to the hands of the male individuals from siphon and faucet surfaces. Further experimentation and research is needed to answer how did they transfer to humans and why they are only seen in males in a very high amount.

When it comes to Bartonella washoensis, we have to urge upon a detail, that is, Bartonella washoensis is pathogenic for humans as it causes meningitis. Somehow, this pathogenic bacteria didn't make its way into the body and cause disease because we know that all the female individuals participated in this experiment are healthy. One possibility is that they are all immunized by meningococcal vaccine. Another interesting detail in our results is that this bacteria is only found in female palm surfaces. We know that this bacteria is transmitted to humans by ground squirrels, so we can argue that one of the female individuals made contact with a ground squirrel that possesses the bacteria and then, just like the previous case, she made her way to the women's restroom in university campus and caused the spread of the bacteria among females. Once again, further experimentation and research is needed to answer why those female individuals didn't have meningitis and why they are only seen in females in a very high concentration.

If we can find out the answers of those questions and detect other bacteria that are gender-specific, it'll be useful in forensic investigations because we can find out if the suspect is a male or female by analying the bacteria genera and determining the gender-specific ones. 


